Sequence data processing

### 2022-04-26 Manuscript/manuscript.Rmd
Cleanup directory for publication

load(file = '../Output/response_model.Rdata')
load(file = '../Output/herit_model.Rdata')

load(file = '../Output/richness_model_passage.Rdata')
load(file = '../Output/richness_model_treat5.Rdata')
load(file = '../Output/richness_model_ch4.Rdata')
load(file = '../Output/phyloseq.RData')
load(file = '../Output/beta_model.Rdata')
load(file = '../Output/da_corncob.Rdata')
load(file = '../Output/dt_data.Rdata')

>>>>

# CH4 models
load(file = '../Output/response_model.RData')
load(file = '../Output/herit_model.RData')

# 16S models
load(file = '../Output/richness_models.RData')
load(file = '../Output/phyloseq.RData')
load(file = '../Output/beta_model.RData')
load(file = '../Output/da_corncob.RData')

### 2022-04-26 R/01-process-flux.Rmd
Cleanup directory for publication

This script processes the standard curve data from the GC on each sample date. It then converts area to methane concentration using the standard curves for every sample. Finally, it computes fluxes (change in concentration over time) as a first-order exponential decay function.

@@

@@

This step first imports the values for the standard curves then calculates the slope of Methane on Area for each sample date.

@@

# These are the final slopes for each Methane~Area standard curve 
# model for each date
sc_ch4 <- 
  sc_ch4 %>% 
  left_join(sc_dates, by = "sc_date") %>% 
  select(flux_date:sc_date, t, r.squared:slope)
```

@@

This step turns the time point data to time intervals in units of days.

@@

```{r compute_time_intervals}

time_data[, c('t0', 't1', 't2', 't3', 't4', 't5')] <- 
  lapply(time_data[, c('t0', 't1', 't2', 't3', 't4', 't5')],
         ymd_hms)

as.days <- function(start, end) {as.numeric(interval(start, end))/60/60/24}

time_data[, c('t1', 't2', 't3', 't4', 't5')] <- 
  lapply(time_data[, c('t1', 't2', 't3', 't4', 't5')],
         function(x) as.days(time_data$t1, x))
time_data <- 
 time_data %>% 
  select(-t0) %>% 
  gather(t, days, t1:t5) %>% 
  mutate(flux_date = factor(flux_date))

@@

This step takes the raw gas chromatograph area data and calculates methane concentration for each time point in units of ppm.

@@

conc_data %>% 
  select(flux_date, jar, t1, t4) %>% 
  mutate(frac_cons = t4/t1) %>% 
  mutate(treat = substring(jar,1,1)) %>% 
  ggplot(aes(x = flux_date, y = frac_cons, color = treat)) +
  geom_jitter(width = 0.2) +
  geom_hline(yintercept = 0.90) +
  geom_hline(yintercept = 0.05)

@@

flux_data <-
  conc_data %>% 
  gather(t, ppm, t1:t5) %>% 
  left_join(time_data, by = c('flux_date', 'jar', 't'))

flux_data <- 
  flux_data %>% 
  mutate(flux_date = factor(ymd(flux_date)), 
         jar = factor(jar),
         t = factor(t)) %>% 
  mutate(rep = factor(substr(jar, 2, 3)),
         treat = factor(substr(jar, 1, 1)))

@@

# Identify the selected jars in the fluxes dataframe
selected$selected <- TRUE
fluxes <- 
  fluxes %>% 
  left_join(selected, by = c('passage', 'jar')) %>% 
  mutate(selected = !is.na(selected)) %>% 
  select(-jar) %>% 
  select(flux_date, passage, treat, rep, ch4:selected)
```

>>>>

This script processes the standard curve data from the gas chromatograph on
each sample date. It then converts area-under-the-curve to methane
concentration using the standard curves for every sample. Finally, it
computes fluxes (change in concentration over time) as a first-order
exponential decay function.

@@

where `flux_date` is the date flux was measured, `treat` is the selection
treatment, either `p` Positive or `n` Neutral, and `n` is the number of jars
for that treatment and sample date. Dates with fewer than 12 samples were due
to broken jars.

@@

This step first imports the values for the standard curves then calculates the slope of Methane on Area for each sample date. On one flux date, a new standard curve was established on the second day of measurement due to drift in the check standards so t1 and t2 had a different standard curve than t3-t5. `flux_date` indicates the main sample data while `sc_date` indicates which standard curve was used on that day.

@@

sc_ch4 %>% 
	kable()

# These are the final slopes for each Methane~Area standard curve 
# model for each date and time point
sc_ch4 <- 
  sc_ch4 %>% 
  left_join(sc_dates, by = "sc_date") %>% 
  select(flux_date:sc_date, t, r.squared:slope)

@@

Remove Passage 6 as explained earlier.

@@

Converts raw data and time time interval in units `days`. The first time point, t0 is when the jars were capped and is ignored because t1 is the first sample injected into the gas chromatograph and so is the first time point with concentrations. Decay constants are calculated between t1 and t5.

```{r compute_time_intervals}
# Standardize time format
time_data[, c('t0', 't1', 't2', 't3', 't4', 't5')] <- 
  lapply(time_data[, c('t0', 't1', 't2', 't3', 't4', 't5')], ymd_hms)

# Convert time to 'days'
as.days <- function(start, end) {as.numeric(interval(start, end))/60/60/24}

time_data[, c('t1', 't2', 't3', 't4', 't5')] <- 
  lapply(time_data[, c('t1', 't2', 't3', 't4', 't5')],
         function(x) as.days(time_data$t1, x))

time_data <- 
 time_data %>% 
  select(-t0) %>% 
  gather(t, days, t1:t5) %>% 
  mutate(flux_date = factor(flux_date))

@@

This step takes the raw gas chromatograph area data and calculates methane concentration for each time point in units of ppm using the standard curves calculated above.

@@

calc_frac_remain <- function(x) {
  if (is.na(x[3])) {
    x[2]/x[1]
  }
  else {
    x[3]/x[1]
  }
}

conc_data %>% 
  select(flux_date, jar, t1, t4, t5) %>% 
  mutate(frac_remain = apply(conc_data[,c('t1', 't4', 't5')], 1, calc_frac_remain)) %>% 
  mutate(treat = substring(jar,1,1)) %>% 
  ggplot(aes(x = flux_date, y = frac_remain, color = treat)) +
  geom_jitter(width = 0.2) +
  geom_hline(yintercept = 0.90) +
  geom_hline(yintercept = 0.05) +
  labs(x = "Flux Date", y = "Fraction of Methane Remaining")

@@

flux_data <-
  conc_data %>% 
  gather(t, ppm, t1:t5) %>% 
  left_join(time_data, by = c('flux_date', 'jar', 't')) %>% 
  mutate(flux_date = factor(ymd(flux_date)), 
         jar = factor(jar),
         t = factor(t)) %>% 
  mutate(rep = factor(substr(jar, 2, 3)),
         treat = factor(substr(jar, 1, 1)))

@@

# Identify the selected jars in the fluxes data frame
selected$selected <- TRUE
fluxes <- 
  fluxes %>% 
  left_join(selected, by = c('passage', 'jar')) %>% 
  mutate(selected = !is.na(selected)) %>% 
  select(-jar) %>% 
  select(flux_date, passage, treat, rep, ch4:selected)
```

### 2022-04-26 R/03-response.Rmd
Cleanup directory for publication

```{r save_model}
save(response_model, compare_response_mods, per_slope, file = '../Output/response_model.Rdata')
```

@@

## Previous Method

Previously, I was using the `metafor` package to test the slope of the difference between the positive and neutral line. For this model, I calculate the difference between the Positive and Neutral treatment and fit that difference as the response. I also calculate the standard error of the difference and included that variance in the model. I then fit the model `difference ~ passage` to test the slope of the difference over time.

```{r eval=FALSE}

# Standard error of the difference
sed <- function(p, n, na.rm = FALSE) {
  sqrt(se(p, na.rm = na.rm) ^ 2 + se(n, na.rm = na.rm)^ 2)
}
# Calculate deviance as p - n
deviance <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate), .groups = "drop") %>% 
  pivot_wider(names_from = treat, values_from = mean) %>% 
  mutate(deviance = p - n)

# Calculate variance of deviance as SED
deviance$se <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  mutate(x = 1:nrow(fluxes)) %>% 
  pivot_wider(names_from = treat, values_from = estimate) %>% 
  group_by(passage) %>% 
  summarize(sed = sed(p, n, na.rm = TRUE), .groups = "drop") %>%
  pull(sed)

# Fit deviance model
fit <- rma(deviance ~ passage, se^2, method="FE", data = deviance)

plot(fit)

# Plot deviance
ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit$beta[1], slope = fit$beta[2])

# Take deviance model parameters and export into a data.frame
rma_output <- data.frame(
Estimate = fit$beta,
se = fit$se,
z.value = fit$zval,
p.value = fit$pval,
upper.ci = fit$ci.lb,
lower.ci = fit$ci.ub
)

# Fit deviance model without passage 2
fit_no2 <- 
filter(deviance, passage != 2) %>% 
rma(deviance ~ passage, se^2, method="FE", data = .)

# Take deviance model withou passage 2 parameters and export into a data.frame

rma_output_no2 <- data.frame(
Estimate = fit_no2$beta,
se = fit_no2$se,
z.value = fit_no2$zval,
p.value = fit_no2$pval,
upper.ci = fit_no2$ci.lb,
lower.ci = fit_no2$ci.ub
)

# R (response to selection) for the model w/o sample 2
R_no2 <- rma_output_no2[[2, 1]]
R <- rma_output[[2, 1]]

# Plot the no2 model
filter(deviance, passage != 2) %>% 
ggplot(., aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit_no2$beta[1], slope = fit_no2$beta[2])

#write_tsv(deviance, '../Output/deviance.tsv')
#saveRDS(rma_output, '../Output/response.rds')
#saveRDS(fit, '../Output/dev_fit.rds')

# Calculate deviance as percentage of p - n
deviance <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  group_by(passage, treat) %>% 
  summarize(mean = mean(estimate), .groups = "drop") %>% 
  pivot_wider(names_from = treat, values_from = mean) %>% 
  mutate(deviance = (p - n) / n)

# Calculate variance of deviance as SED
deviance$se <-
  fluxes %>% 
  select(treat, passage, estimate) %>% 
  mutate(x = 1:nrow(fluxes)) %>% 
  pivot_wider(names_from = treat, values_from = estimate) %>% 
  group_by(passage) %>% 
  summarize(sed = sed(p, n, na.rm = TRUE), .groups = "drop") %>%
  pull(sed)

# Fit deviance model
fit <- rma(deviance ~ passage, se^2, method="FE", data = deviance)

plot(fit)

# Plot deviance
ggplot(deviance, aes(x = passage, y = deviance, ymin = deviance - se, ymax = deviance + se)) + 
  geom_pointrange(color = 'darkorange2') + 
  geom_abline(intercept = fit$beta[1], slope = fit$beta[2])

# Take deviance model parameters and export into a data.frame
rma_output <- data.frame(
Estimate = fit$beta,
se = fit$se,
z.value = fit$zval,
p.value = fit$pval,
upper.ci = fit$ci.lb,
lower.ci = fit$ci.ub
)

# Fit separte slopes for Positive and Neutral with robust M-estimator regression

library(MASS)
anova(
fit <- rlm(estimate ~ passage * treat, data = fluxes),
fit <- rlm(estimate ~ passage + treat, data = fluxes))

summary(fit)
anova(fit)
new_data <- cbind(fluxes, predict(fit, interval = 'confidence'))
colnames(new_data) 
ggplot(new_data, aes(x = passage, y = estimate, color = treat)) +
  theme_bw() +
  geom_jitter() +
  geom_line(aes(passage, fit)) +
  geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)

ggplot(new_data, aes(x = passage, y = estimate)) +
  theme_bw() +
  geom_jitter(aes(color = treat)) +
  geom_line(aes(passage, fit, color = treat)) +
  geom_ribbon(aes(ymin=lwr,ymax=upr, group = treat), alpha=0.3) +
  ylim(-0.01, 0.08)
ggplot(fluxes, aes(x = passage, y = estimate, color = treat)) + 
  geom_point() + 
  stat_smooth(method='lm')
#ggsave('raw_fluxes.pdf')
```

>>>>

```{r save_model}
save(response_model, compare_response_mods, per_slope, file = '../Output/response_model.RData')
```

@@

### 2022-04-26 R/04-heritability.Rmd
Cleanup directory for publication

This script imports the heritability data. This dataset includes the mean flux
withn treatmentxgeneration for 1) all parents, 2) selected parents, and 3)
offspring. Then, it plots offspring on the selection parents with the two
treatments combined and then separately for the two treatments.

@@

The negative heritability of methane oxidation rate in the Neutral selection treatment may result from selection on alternative characters in those populations. The strong, positive heritability in the Positive treatment ... . It seems strange that the

@@

```{r save_heritability}
save(herit_model, file = '../Output/herit_model.Rdata')
```

######################################################################
# End of the way I should be doing it
################################################################

The Old way:

It then calculates S, the selection differential, as the mean difference between the selected parents and all parents. Because passage 2 was accidentaly selected for the lowest methane oxidation rate, I take the absolute value of this difference, i.e. |S|. Next, it imports the deviance data and calculates R, the respones to selection, as the slope of mean difference in flux (positive - neutral) over time.

An alternative approach would be to calculate R directly instead of approximating it using the slope as in the equation offspring - parent. In this case, I would not be using the deviance, but instead the raw fluxes. I can do that with the heritability dataset.

# Function for the Breeder's equation

```{r eval=FALSE, include=FALSE}
breeders <- function(R, S) {
  R / S
}

#

selected <- 
  fluxes[paste0(fluxes$passage, fluxes$jar) %in% paste0(selected_jars$passage, selected_jars$jar), ] %>%
  select(treat, log_ch4, passage) %>% 
  group_by(treat, passage) %>% 
  summarize(selected = mean(log_ch4), .groups = 'drop')

parents <- 
  fluxes %>%
  select(treat, log_ch4, passage) %>% 
  group_by(treat, passage) %>% 
  summarize(parental = mean(log_ch4), .groups = 'drop')

offspring <- 
  fluxes %>% 
  select(passage, treat, log_ch4) %>% 
  filter(passage != 1) %>% 
  mutate(passage = passage - 1) %>% 
  group_by(treat, passage) %>% 
  summarize(offspring = mean(log_ch4), .groups = 'drop')

heritability <- 
  offspring %>% 
  left_join(selected, by = c('passage', 'treat')) %>% 
  left_join(parents, by = c('passage', 'treat'))

# Matt's approach: h2 for each generation
h2_per_gen  <- 
heritability %>% 
  mutate(R = offspring - parental) %>% 
  mutate(S = selected - parental) %>% 
  mutate(h2 = breeders(R, S))

write_tsv(h2_per_gen, 'Output/h2_per_gen.tsv')

ggplot(h2_per_gen, aes(x = passage, color = treat, y = h2)) +
  geom_point()

h2_per_gen %>% 
  select(treat, passage, h2)

h2_per_gen %>% 
  select(treat, passage, h2) %>% 
  group_by(treat) %>% 
  summarize(mean_h2 = mean(h2), sd_h2 = sd(h2), .groups = 'drop')

# Calculate R using regression of deviance over time.
# Calculate S using math.

dev_fit <- readRDS('../Output/dev_fit.rds')
R <- dev_fit$beta[2]

S <-
  heritability %>%
  filter(treat == "p") %>%
  mutate(S = abs(selected - parental)) %>%
  summarize(S = mean(S))  %>% 
  pull()

h2_reg_dev <- breeders(R, S)

lm_fit <- 
  fluxes %>% 
  filter(treat == 'p') %>% 
  select(treat, passage, ch4) %>% 
  lm(ch4 ~ passage, data = .)

R_not_dev <- pull(tidy(lm_fit)[2, 2])
h2_reg_abs <- breeders(R_not_dev, S)

# Calculate R as regression of offspring on parents

ggplot(heritability, aes(x = parental, y = offspring)) +
  geom_point() + 
  stat_smooth(method = 'lm', se = FALSE)
summary(lm(offspring ~ parental, data = heritability, subset = treat == 'p'))
# Calculate h2 (heritability) using math, not regression
# For each generation and then calculate the mean/se
h2_calc_method <-
  heritability %>%
  filter(treat == "p") %>%
  mutate(S = abs(selected - parental)) %>%
  mutate(R = offspring - parental) %>%
  mutate(h2 = breeders(R, S)) %>%
  summarize(S = mean(S),
            R = mean(R),
            se = se(h2),
            h2 = mean(h2))

h2_calc_method_no_2 <-
  heritability %>%
  filter(treat == "p") %>%  
  filter(passage != 2) %>%
  mutate(S = abs(selected - parental)) %>%
  mutate(R = offspring - parental) %>%
  mutate(h2 = breeders(R, S)) %>%
  summarize(S = mean(S),
            R = mean(R),
            se = se(h2),
            h2 = mean(h2))

h2 <- 
  h2_calc_method %>% 
  mutate(method = "calc p")

h2_no2 <- 
  h2_calc_method_no_2 %>% 
  mutate(method = "calc p no 2")

h2 <- rbind(h2, h2_no2)

# Cumulative response to selection

parent <- heritability[[1, 3]]
offspr <- heritability[[8, 4]]
select <- heritability[[8, 5]]
R = offspr - parent
S = select - parent
h2 = breeders(R, S)

heritability[,'cum_res'] <- NA
heritability$cum_sel <- NA
herit_p <- filter(heritability, treat == 'p')
for (i in seq(nrow(herit_p))) {
  off <- sum(herit_p$offspring[1:i])
  par <- sum(herit_p$parental[1:i])
  sel <- sum(herit_p$selected[1:i])
  herit_p$cum_res[i] <- off - par
  herit_p$cum_sel[i] <- sel - par
}
herit_p$cum_h2 <- herit_p$cum_res / herit_p$cum_sel

ggplot(herit_p, aes(x = cum_sel, y = cum_res)) + 
  theme_classic() +
  labs(x = 'Cumulative selection differential',
       y = 'Cumulative response to selection') +
  geom_point() + 
  stat_smooth(method = 'lm', se = FALSE)
summary(lm(cum_res ~ cum_sel, data = herit_p))

# Regression of offspring on mid-parent

selected_jars <- read_csv('../Data/selected.csv')
fluxes <- read_tsv('../Output/fluxes.tsv')

# Pull out the selected parental jars
selected <- 
  fluxes[paste0(fluxes$passage, fluxes$jar) %in% paste0(selected_jars$passage, selected_jars$jar), ] %>%
  select(treat, ch4, passage) %>% 
  group_by(treat, passage) %>% 
  summarize(parental = mean(ch4), .groups = 'drop')

offspring <- 
  fluxes %>% 
  select(passage, treat, ch4) %>% 
  filter(passage != 1) %>% 
  mutate(passage = passage - 1) %>% 
  rename(offspring = ch4)

heritability <- 
  offspring %>% 
  left_join(selected, by = c('passage', 'treat'))

heritability

ggplot(heritability, aes(parental, offspring)) +
  theme_classic() +
  geom_point() + 
  stat_smooth(method = 'lm', se = FALSE)

summary(lm(offspring ~ parental, data = heritability))

ggplot(heritability, aes(parental, offspring, color = treat)) +
  theme_classic() +
  geom_point() + 
  stat_smooth(method = 'lm', se = FALSE)
summary(lm(offspring ~ parental * treat, data = heritability))
```

>>>>

This script imports the heritability data. This dataset includes the mean flux
within treatment by generation for 1) all parents, 2) selected parents, and 3)
offspring. Then, it plots offspring on the selected parents with the two
treatments combined and then separately for the two treatments.

@@

@@

```{r save_heritability}
save(herit_model, file = '../Output/herit_model.RData')
```

### 2022-04-26 R/07-generate-physeq.Rmd
Cleanup directory for publication

```{r import_data}
load(file = "Output/dada2_output.RData")

@@

# Create final phyloseq

@@

# 
# 
# seqtab.nochim.filtered <-
#   seqtab.nochim[rownames(seqtab.nochim) %in% rownames(fluxes_filtered),]
# 
# ps <- phyloseq(otu_table(seqtab.nochim.filtered, taxa_are_rows=FALSE), 
#                                      sample_data(fluxes_filtered),
# 
#                tax_table(taxa))
# 
# 
# 
# dna <- Biostrings::DNAStringSet(taxa_names(ps))
# names(dna) <- taxa_names(ps)
# ps <- merge_phyloseq(ps, dna)
# taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))

>>>>

```{r import_data}
load(file = "HPC_Output/dada2_output.RData")

@@

@@
